Mergeable Types

### 2017-02-25 paper/mergeable-types.tex
edits

Merge functions for simple data types are easy to write. For instance,
the merge logic for counter is simple enough that it does not require
great effort to identify the $P$ and $Q$ that leda to the
concurrent counter versions, and transform them into $P'$ and $Q'$
such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
sophisticated data types, this process is more involved, and may require
the programmer to perform non-trivial reasoning. We now present an
alternative definition of a mergeable type that is stronger than
Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
non-trivial merge functions.

\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type-2}
A type $t$ is said to be mergeable if and only if the following
conditions hold:
\begin{itemize}
  \item There exists a function $W: t \times t \rightarrow (t
  \rightarrow t)\; \C{list}$ that accepts an object $A:t$ and its
  semantic successor $B:t$, and returns a minimal sequence of
  primitive morphisms $P^*$, whose composition maps $A$ to $B$. That
  is, $W(A,B) = P^*$, where $\denot{P^*}(A) = B$, and there does not
  exist an $R^*$ such that $\denot{R^*}(A) = B$ and $\length{R^*} <
  \length{P^*}$

  \item There exists a function $M: (t \rightarrow t)^*\!\times\!(t
  \rightarrow t)^* \;\rightarrow\; (t \rightarrow t)\!\times\!(t
  \rightarrow t)$ that accepts a pair of minimal sequences of
  primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
  $Q'$ and $P'$, such that for any object $A:t$, $(Q' \circ
  \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
\end{itemize}
\end{definition}

\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (a string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:mergeable-string-eg} shows the edit script ...

>>>>

Merge functions for simple data types are easy to write. For instance,
the merge logic for counter is simple enough that it does not require
great effort to identify the $P$ and $Q$ that leda to the
concurrent counter versions, and transform them into $P'$ and $Q'$
such that the diagram in Fig.~\ref{fig:pushouts} commutes. For more
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.35]{Figures/pushouts-2}
\caption{Commutative diagram for Def.~\ref{def:mergeable-type-2}.}
\label{fig:pushouts-2}
\end{wrapfigure}
sophisticated data types, this process is more involved, and may require
the programmer to perform non-trivial reasoning. We now present an
alternative definition of a mergeable type that is stronger than
Def.~\ref{def:mergeable-type}, but also serves as a recipe to build
non-trivial merge functions.

\paragraph{Notation} As mentioned earlier, we distinguish between the
named functions defined by a data type library (primitive morphisms),
and their compositions (morphisms). A sequence (list) of primitive
morphisms is written $P^*$ or $Q^*$. The composition of a sequence of
primitive morphisms (written $\denot{P^*}$) is their right-to-left
composition, i.e., $\C{fold\_left}\;(\lambda f.\lambda g. f \circ
g)\; id\; P^*$, where $\C{fold\_left}$ has its usual OCaml type. The
length of a sequence $S$ is written $\length{S}$.

\begin{definition} [\bfseries Mergeable Type]
\label{def:mergeable-type-2}
A type $t$ is said to be mergeable if and only if the following
conditions hold:
\begin{itemize}
  \item There exists a function $W: t \times t \rightarrow (t
  \rightarrow t)^*$ that accepts an object $A:t$ and its
  semantic successor $B:t$, and returns a minimal sequence of
  primitive morphisms $P^*$, whose composition maps $A$ to $B$. That
  is, $W(A,B) = P^*$, where $\denot{P^*}(A) = B$, and there does not
  exist an $R^*$ such that $\denot{R^*}(A) = B$ and $\length{R^*} <
  \length{P^*}$

  \item There exists a function $M: (t \rightarrow t)^*\!\times\!(t
  \rightarrow t)^* \;\rightarrow\; (t \rightarrow t)\!\times\!(t
  \rightarrow t)$ that accepts a pair of minimal sequences of
  primitive morphisms, $P^*$ and $Q^*$, and returns a pair of morphisms,
  $P'$ and $Q'$, such that for any object $A:t$, $(Q' \circ
  \denot{P^*})(A) = (P' \circ \denot{Q^*})(A)$.  
\end{itemize}
Commutativity diagram of Fig.~\ref{fig:pushouts-2} visualizes this
definition. 
\end{definition}

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which computes the \emph{edit
distance} between an object $A:t$ and its semantic successor $B:t$.
The sequence of primitive morphisms it returns ($P^*$) is called an
\emph{edit script}. Replaying the edit script on $A$ produces $B$
(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an \emph{operational
transformation} of the primitive morphisms in an edit script. An
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.4]{Figures/string-eg}
\caption{Merging strings with edit scripts and operational
transformations}
\label{fig:string-eg}
\end{wrapfigure}
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$. We will now consider examples of mergeable types that reify the
notions of edit scripts and operational transformations.
\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (a string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:string-eg} illustrates.
